Statistical methods for relative quantification of post-translational modifications in global proteomics experiments

### 2021-09-23 main/MSstatsPTM_Main.tex
Updated results section and moved additional considerations to main

Further results for this experiment are provided in Supplementary \todo{link to Sec. S4.2}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Re-analysis of published datasets}
\subsubsection*{IpaH7.8}

\cite{LUCHETTI2021}

\subsubsection{Shigella flexneri}
To evaluate our approach we reanalyzed an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri). \cite{Maculins} Tandem mass tagging was used to quantify changes in total proteins, phosphorylation, and ubiquitination in Wild type (WT) controls and ATG16L1-deficient (cKO) bone marrow macrophages (BMDMs) either infected or uninfected with Shigella flexneri. 11-plex isobaric multiplex was used with two separate mixtures. Cell lysates were prepared over three different time periods, either uninfected, infected at early (45-60 min) or infected at late (3-3.5 hour) time points. There were 103,700 peptides mapping to 9,430 proteins in the global profiling run. Additionally, there were 25,600 unique phosphorylation sites and 12,400 KGG ubiquitination sites. About ~90\% of the identified modified peptides derived from proteins that were also quantified in the global profiling run.

The major results are summarized below.

\subsubsection{USP30}

@@

We proposed a general statistical modeling framework for PTM characterization. The framework is designed for bottom-up MS workflows, which are characterized with variations from multiple convoluted sources, frequent missing data, and associated uncertainty in the conclusions. The framework is general and is applicable to a variety of experimental designs. It outperforms the ad-hoc methods underlying the t-test, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, batch effects, and different acquisition methods. The framework allows us to plan for subsequent experiments, and choose the appropriate number of replicates in consideration of adjustment with respect to protein abundance.

Our results show that when measurements from multiple related conditions are available, the proposed approach for joint modeling and summarization of all the LC-MS/MS runs leads to more sensitive PTM significance analysis and more accurate and precise quantification than when separately analyzing conditions of runs. The gain is due to a more efficient use of the data, and to a more accurate understanding of the systematic and random variations. The proposed framework can be extended beyond the experimental designs with multiple batches and conditions discussed above. For example, it can represent experimental designs with even more complex structures, such as time series or factorial investigations.

@@

>>>>

Further results of this experiment are provided in Supplementary \todo{link to Sec. S4.2}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Re-analysis of published datasets}

\subsubsection*{IpaH7.8}

The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD).\cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells. Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The experiment was inbalanced with two bioreplicates per condition for all conditions except for infected 1 hour. About ~95\% of the identified modified peptides derived from proteins that were quantified in the global profiling run.

Looking at modified peptide $GSDMD\_HUMAN|P57764$ we can see the advantage of the proposed method. The modified peptide shows a flat abundance change between the infected 1 hour, 4 hour, and 6 hour conditions. This is contrasted with a strong negative change in the global profiling run between the same conditions. Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted pvalue was insignificant at $.0644$. After adjusting for changes in the global protein abundance, the fold change is much higher, $2.79$, and the adjusted pvalue became very significant, $5.25e^{-8}$. In this case the effect of the modified peptide is strongly confounded with changes in the global protein. The proposed method allows us to remove this confounding and estimate the true effect. The results of this can be seen in Figure \todo{add profile plot}.

Further details and results of this experiment are provided in Supplementary \todo{link to Sec. S5.1}.

\subsubsection*{Shigella flexneri}
To evaluate our approach we reanalyzed an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri). \cite{Maculins} Multiplex proteomics was used to quantify the abundance of total protein, phosphorylation, and ubiquitination in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri. The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours). Quantifying the total protein along with the post-translational modifications allowed us to adjust for changes in total protein and see the true impact of the site specific modifications. Two mixtures using 11-plex were ran over the six conditions. The six conditions were split between 11 channels leading to the experimental design being unbalanced. Each mixture contained two replicates per early and late WT and KO conditions. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT. About ~90\% of the identified modified peptides derived from proteins that were also quantified in the global profiling run.

Two specific modified peptides, $TTP\_MOUSE|P22893_S178$ and $KI67\_MOUSE|E9PVX6_T215$, show the power of the proposed method.  The profile plot of modified peptide $TTP\_MOUSE|P22893_S178$ shows a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions. However, the global profiling run shows a similar log fold change of $2.015$ between  the same conditions. This indicates that the abundance change in the modified peptide is nearly entirely due to changes in the global protein. When adjusting for the global protein the modified peptide's adjusted pvalue becomes insignificant, going from $.0009$ to $.275$. In contrast $KI67\_MOUSE|E9PVX6_T215$ originally shows a small positive log fold change of $.212$ between conditions WT\_Early and WT\_Uninfected. This is contrast with a negative fold change in the global profiling run of $-.616$. In adjusting the modified peptide for changes in the global protein, the log fold change becomes .828 and the adjusted pvalue goes from $.452$ to $.037$.

Further details and results of this experiment are provided in Supplementary \todo{link to Sec. S5.2}.

\subsubsection*{USP30}

This experiment looked into the relationship between USP30 and protein kinase PINK1, and their association with Parkinson’s Disease. Ubiquitination site profiling was performed and the modified site abundance was analyzed. Four conditions were tested with two biological replicates per condition. The conditions were as follows: CCCP, USP30 overexpression (USP30 OE), Combo, and Control. Label-free mass spectrometry quantification was used to quantify the abundance of modified peptides. A corresponding mixed effects model was fit per modification and global protein as described previously in this supplementary.

In contrast to other experiments analyzed in this paper, there was no unmodified global protein profiling run performed in this experiment. Once identification and quantification of the Ubiquitinated profiling was performed, peptides which were unmodified were extracted and used in place of a global profiling run. This resulted in a significant lack of overlap between modified and unmodified peptides. Any modified peptide without a corresponding unmodified protein could not be adjusted. Of the 10,799 modified peptides identified, only 4526 had a corresponding unmodified run and could be adjusted. Finally, using this method also resulted in a low feature count for the unmodified protein model.

Given the lack of matches in the global profiling run, the proposed method was unable to be applied to as many modified peptides as in the previous examples. However we can still see the power of the method when looking at peptide $P52209\_K059$. The abundance between conditions appears flat in the modified peptide profile plot, however this is contrast with a large negative change in the corresponding global protein. Using the proposed method we get increased information that would otherwise have been missed.

Further details and results of this experiment are provided in Supplementary \todo{link to Sec. S5.3}.

@@

We proposed a general statistical modeling framework for PTM characterization. The framework is designed for bottom-up MS workflows, which are characterized with variations from multiple convoluted sources, frequent missing data, and associated uncertainty in the conclusions. The framework is general and is applicable to a variety of experimental designs. It outperforms the ad-hoc methods underlying t-test and limma, and yields accurate results in the broad type of experimental circumstances, including the presence of missing values, changes in protein abundance, few representative peptides, and different acquisition methods. The framework allows us to plan for subsequent experiments, and choose the appropriate number of replicates in consideration of adjustment with respect to protein abundance.

Our results show that when measurements from multiple related conditions are available, the proposed approach for joint modeling and summarization of all the LC-MS/MS runs leads to more sensitive PTM significance analysis and more accurate and precise quantification than when separately analyzing conditions of runs. The gain is due to a more efficient use of the data, and to a more accurate understanding of the systematic and random variations. The proposed framework can be extended beyond the experimental designs with variation from multiple sources discussed above. For example, it can represent experimental designs with even more complex structures, such as time series or factorial investigations.

@@

Additionally, expression levels of PTM sites can be convoluted with each other if there are two or more modification sites per peptide. In the current implementation the effect of a specific modification in a peptide with multiple modifications cannot be quantified. One potential solution to this is to measure the abundance of peptides with one modification and use this to adjust the peptide with multiple sites to remove the convolution. However, this method would most likely run into major challenges due to sparsity of features for both modified peptides with a single and multiple modification sites, as well as a potential lack of matching single site modified peptides. A more complex approach to addressing this problem is most likely necessary.

### 2021-09-23 supplementary/ptm_sm.tex
Updated results section and moved additional considerations to main

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage

\todo{move to main manuscript}

\section{Further Considerations}

\subsubsection{Multiple sites per protein}

Expression levels of PTM sites are adjusted based on the abundance of their originating protein. Since the same reference is used for all sites, it introduces correlation among estimates and test statistics for those sites. This may cause issues in controlling false discovery rate (FDR).
We investigated the property by simulating data of two conditions and 1000 proteins with different numbers of PTM sites and comparing the results of the proposed approach and the linear model with no adjustment. 
%We investigated the property by simulating data of two conditions and 1000 proteins with different numbers of PTM sites. 
A fraction (50\%) of the 1000 proteins had no changes between conditions while systematic changes were simulated for the rest of the proteins. Multiple testing correction was performed using the Benjamini and Hochberg's method. 
Performances of the considered approaches were assessed by their actual FDR, calculated as the fraction of proteins with adjusted $p$-values $<0.05$ among the proteins with true differences. The results are summarized in \sfigref{fig:prot_fdr}.
%Performances of the considered approaches were assessed by their power and FDR. The power was calculated as the fraction of proteins with adjusted $p$-values $<0.05$ among those proteins with true differences between conditions, while the FDR was calculated as the fraction of proteins with adjusted $p$-values $<0.05$ among the proteins with true differences. The results are summarized in \sfigref{fig:prot_fdr}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\section*{References}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\section*{References}
